sampbias, a method for quantifying geographic sampling biases in species distribution data

### 2020-07-30 appendix_S1.Rmd
change plotting and adapt simulations

```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262, 2016, https://doi.org/10.15468/dl.7fg4zx), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis."}
#libraries
library(raster)
library(ggplot2)
library(magrittr)
library(sp)
library(dplyr)
library(viridis)
library(readr)
library(sf)
library(geosphere)
library(tidyr)
library(forcats)

@@

born2 <- st_read("empirical_analyses/Borneo.kml")
born2 <- sf:::st_zm(born2$geom)
born2 <- as(born2, 'Spatial')

@@

born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
born2 <- sf:::st_zm(born2$geom)
born2 <- as(born2, 'Spatial')
born <- raster::intersect(landmass, born2)

@@

dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
  mutate(records = ifelse(type=="empirical", rar * nrow(occ), rar * 100000)) %>% 
  mutate(records_area = round(records / area * 10000, 2)) %>% 
  select(contains("w"), ID, type, records_area) %>% 
    pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias))
```

@@

```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate."}

@@

ggplot(data = plo)+
  geom_boxplot(aes(x = bias, 
                   y = posterior_estimate,
                 fill = as.factor(records_area)))+
  ylab("Posterior weight")+
  xlab("Bias factor")+
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  theme_bw()
```

>>>>

```{r, fig.height = 8, fig.cap="The example dataset of mammal occurrences from the island of Borneo, as downloaded from www.gbif.org (n = 6,262, 2016, https://doi.org/10.15468/dl.7fg4zx), and the geographic gazetters of main cities, roads, rivers and airports used for the sampbias analysis. Note that the records in the sea around Borneo were excluded by with the help of the user-provided study area."}
#libraries
library(raster)
library(ggplot2)
library(magrittr)
library(sp)
library(dplyr)
library(viridis)
library(readr)
library(sf)
library(geosphere)
library(tidyr)
library(forcats)

@@

born2 <- st_read("empirical_analyses/Borneo.kml", quiet = TRUE)
born2 <- sf:::st_zm(born2$geom)
born2 <- as(born2, 'Spatial')

@@

born <- raster::intersect(landmass, born2)

@@

dat <- read_csv("empirical_analyses/simulations/weight_estimates.csv") %>% 
  mutate(records_area = round(rar / area * 10000, 2)) %>% 
  select(contains("w"), ID, type, records_area) %>% 
    pivot_longer(cols = contains("w_"), names_to = "bias",
                 values_to = "posterior_estimate") %>%
    mutate(bias = gsub("w_", "", .data$bias))
```

@@

```{r, out.width = "6in", fig.cap = "Projected bias surface for the distribution data of mammals from Borneo (www.gbif.org, 2016, https://doi.org/10.15468/dl.7fg4zx), at different levels of downsampling, from a random replicate. The plotted extent is adapted to the study area, which may not cover the entire island for rarified datasets with few records (A) and B))"}

@@

```{r, echo = FALSE, fig.height = 8, fig.cap="The bias weights ($w$) defining the effects of each bias factor estimated for the smallest simulated dataset without bias."}

ggplot(data = plo)+
  geom_histogram(aes(posterior_estimate, fill = as.factor(records_area)), bins = 50) +
  scale_fill_viridis(discrete = TRUE, name = "Records per\n10,000 sqkm")+
  facet_grid(rows = vars(records_area), cols = vars(bias))+
  scale_x_continuous(breaks = c(0,0.003, 0.006, 0.009), limits = c(0,0.01))+
  xlab("Posterior weight")+
  ylab("Count")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(),
        legend.position = "none")

### 2020-07-30 manuscript.rmd

## Distance calculation
*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area. \textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g. rainforest) or an area of special interest (e.g. a national park)}. For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$. The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 110 km precision at the equator.  Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global dataset would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c. 100 -- 10,000 m). \textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid---may be provided by the user.}

@@

To quantify the amount of bias associated with each factor, we jointly estimate the parameters $q$ and $\mathbf{w}$ in a Bayesian framework. We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
$$
P(q, \mathbf{w} | \mathbf{S}) \propto \prod_{i=1}^{N}{ Poi(S_i | \lambda_i) } \times P(q) P(\mathbf{w}) \ \ \ \ (2)
$$
where the likelihood of sampled occurrences $S_i$ within each cell $Poi(S_i | \lambda_i)$ is the probability mass function of a Poisson distribution with rate per cell defined as in Eqn. (1). The likelihood is then multiplied across the $N$ cells considered. We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively. \textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$\textcolor{blue}{, all of which must be positive and have support [0, +inf]. We designed the priors to be informative (i.e. not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests. For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3. Custom priors, within the flexible family of gamma distributions are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}

@@

We summarize the results by mapping the estimated sampling rates ($\lambda_i$) across space. These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates. Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups. Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in sampling rates, which are not linked with bias factors. For instance, the unbiased sampling rate $q$ is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases ($\mathbf{w}$) might be similar across the two groups.

@@

*Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage. To test the performence of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals on Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each). The results showed that estimates of} $\mathbf{w}$ \textcolor{blue}{ and the projection of the bias effect in space in a biased dataset were robust to decreases in data, although uncertainty increased (Figs. S2nd S3, in Appendix S2). 2) in a second simulation we  randomly sampled occurrence records across Borneo as an example of an unbiased dataset. We sampled 100,000, 50,000, 25,000, and 10,000 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}. \textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs. S4 and S5, in Appendix S2), however, we note that the analyses took a very long time to converge for datasets with no sampling bias and datasets with less than 10,000 points. In general, the more records, the more precise the parameter estimates will be.}

@@

```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the estimated sampling rates in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org. GBIF.org, 2016). The colours show the projection of the \\textcolor{blue}{log-transformed} sampling rates (i.e. expected number of occurrences per cell) given the inferred \textit{sampbias} model. The highest undersampling is in the centre of the island.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection_adapted.jpg")
```

>>>>

## Distance calculation
*Sampbias* uses gazetteers of the geographic location of bias factors (hereafter indicated with B) to generate a regular grid across the study area. \textcolor{blue}{By default the study area is defined by the geographic extent of the study dataset, but it can also be customized via user-defined polygons, for instance to limit the analyses to an environmentally homogeneous region (e.g. rainforest) or an area of special interest (e.g. a national park). In this case all occurrence records outside the user-defined area will be disregarded for the analysis.}. For each grid cell $i$, we then compute a vector $X_i(j)$ of minimum distances (straight aerial distance, "as the crow flies") to each bias factor $j \in B$. The resolution of the grid defines the precision of the distance estimates, for instance a 1x1 degree raster will yield approximately a 110 km precision at the equator. Due to the assumption of homogeneous sampling and a computational trade-off between the resolution of the regular grid and the extent of the study area (for instance, a 1 second resolution for a global dataset would become computationally prohibitive in most practical cases), *sampbias* is best suited for local or regional datasets at high resolution (c. 100 -- 10,000 m). \textcolor{blue}{Since the differences in grid cell size are negligible on the local and regional scale, sampbias uses a latitude/longitude grid by default, but a custom grid in any projection and coordinate reference system---for instance an equal area grid---may be provided by the user.}

@@

To quantify the amount of bias associated with each factor, we jointly estimate the parameters $q$ and $\mathbf{w}$ in a Bayesian framework. We use Markov Chain Monte Carlo (MCMC) to sample these parameters from their posterior distribution:
$$
P(q, \mathbf{w} | \mathbf{S}) \propto \prod_{i=1}^{N}{ Poi(S_i | \lambda_i) } \times P(q) P(\mathbf{w}) \ \ \ \ (2)
$$
where the likelihood of sampled occurrences $S_i$ within each cell $Poi(S_i | \lambda_i)$ is the probability mass function of a Poisson distribution with rate per cell defined as in Eqn. (1). The likelihood is then multiplied across the $N$ cells considered. We used exponential priors on the parameters $q$ and $\mathbf{w}$, $P(q) \sim \Gamma(1, 0.01)$ and $P(\mathbf{w}) \sim \Gamma(1, 1)$, respectively. \textcolor{blue}{We chose exponential priors because they represent the standard choice for rate parameters such as} $q$ \textcolor{blue}{and the weights} $\mathbf{w}$\textcolor{blue}{, all of which must be positive and have support [0, +inf]. We designed the priors to be informative (i.e. not allowing negative values) and yet vague enough to encompass a much wider range of parameter space than the range of values observed in empirical tests. For instance, the posterior weights estimated in our empirical study ranged from 0 to 0.013, while the prior distribution applied on these parameters, Exp(1) has a} 95% \textcolor{blue}{of it density in range 0 – 3. Custom priors, within the flexible family of gamma distributions, which include the exponential priors used here, are possible via the} `prior_q` and `prior_w` \textcolor{blue}{arguments of the} `calculate_bias` \textcolor{blue}{function.}

@@

We summarize the results by mapping the estimated `textcolor{blue}{(the relative deviation of sampling rate from the maximum
sampling rate (or on user choice the estimated sampling rates}, $\lambda_i$) across space. These rates represent the expected number of sampled occurrences for each grid cell and provide a graphical representation of the spatial variation of sampling rates. Provided that the cells are of equal size, the estimated rates will be comparable across data sets, regions, and taxonomic groups. Analysing different regions, biomes, or taxa in separate analyses allows to account for differences in sampling rates, which are not linked with bias factors. For instance, the unbiased sampling rate $q$ is expected to differ between a highly sampled clade like birds and under-sampled groups of invertebrates, but their sampling biases ($\mathbf{w}$) might be similar across the two groups.

@@

*Sampbias* \textcolor{blue}{is designed to work with sparsely sampled datasets, and to estimate bias effects from datasets with low coverage. To test the performance of} *sampbias* \textcolor{blue}{on small datasets we ran two simulation experiments: 1) we ran the analyses as presented above on the distribution of mammals in Borneo, sub-sampling of the initial dataset (to 6,262, 3,131, 1,566, 626, 62 records for all of Borneo respectively, 5 replicates each). The results showed that parameter estimates and the projection of the bias effect in space were robust to decreases in data, although uncertainty increased (Figs. S2nd S3, in Appendix S2). 2) in a second simulation we  randomly sampled occurrence records across Borneo as an example of an unbiased dataset. We sampled 100,000, 50,000, 25,000, and 10,000 records and ran a} *sampbias* \textcolor{blue}{analysis with the same settings as for the empirical dataset}. \textcolor{blue}{The results show that} *sampbias* \textcolor{blue}{reliably rejected a biasing effect even for small datasets (Figs. S4 and S5, in Appendix S2). In general, the more records, the more precise the parameter estimates will be.}

@@

```{r spatial, out.width = "6in", echo = FALSE, fig.cap = "Spatial projection of the sampling bias in an empirical example dataset of mammal occurrences on the Indonesian island of Borneo (downloaded from www.gbif.org. GBIF.org, 2016). The colours show the projection of the \\textcolor{blue}{the relative deviation of sampling rate from the maximum sampling rate given the inferred} \\textit{sampbias} model. For instance, a value of -25 indicates a drop of 25\\% compared to the highest sampling rate. The highest undersampling is in the centre of the island.", eval = TRUE}
knitr::include_graphics(path = "ms_figures/figure_empirical_results_spatial_projection_adapted.jpg")
```

### 2020-07-30 title_page.Rmd
manuscript changes

---
title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
# title: "A global Red List of orchids supports the value of automated conservation assessment"
author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
output: 
  bookdown::pdf_document2:
  #word_document:
    toc: false
    toc_depth: 1
    number_sections: false
header-includes: |
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias}
    \fancyhead[R]{ }
    \pagenumbering{gobble}
linestretch: 1.3
fontsize: 12pt
---

@@

**Word count** -- Abstract: 3,051 words; Total: 208 words, 2 figures

>>>>

---
title: "sampbias, a method for quantifying geographic sampling biases in species distribution data"
author: "Alexander Zizka^1,2^, Alexandre Antonelli^3,4,5^, Daniele Silvestro^3,4,6^"
output: 
  bookdown::pdf_document2:
  #word_document:
    toc: false
    toc_depth: 1
    number_sections: false
header-includes: |
    \usepackage{fancyhdr}
    \pagestyle{fancy}
    \fancyhead[L]{sampbias}
    \fancyhead[R]{ }
    \pagenumbering{gobble}
linestretch: 1.3
fontsize: 12pt
---

@@

**Word count** -- Abstract: 207 words; Main text (excluding references): 3019 words, 2 figures
